The Proofs

### 2024-03-15 oopsla24/tex/04-algorithm.tex
wip

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  infers a type of \emph{annotated} applicative let binder.
  First, it infers the type of the head of the application,
  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
  After that, it makes a recursive call
  to the application inference procedure,
  returning an algorithmic type $[[↑uQ]]$, 
  that must be instantiated to a subtype of 
  the annotation $[[↑uP]]$.
  Then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
  check whether the instantiation to the annotated type $[[↑iP]]$ is possible,
  and if it is, the algorithm infers the type of the body in the extended context,
  and returns it as the result.

>>>>

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  infers a type of \emph{annotated} applicative let binder.
  First, it infers the type of the head of the application,
  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
  After that, it makes a recursive call
  to the application inference procedure,
  returning an algorithmic type $[[iM']]$, 
  that must be a subtype of the annotation $[[↑uP]]$.

  Then premise $[[Γ; Θ ⊨ iM' ≤ ↑iP ⫤ SC2]]$
  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
  check that $[[iM']]$ can be instantiated to the annotated type $[[↑iP]]$,
  and if it is, the algorithm infers the type of the body in the extended context,
  and returns it as the result.

### 2024-03-15 proofs/main.tex
wip

@@

Let us discuss the selected rules of the declarative system:
\begin{itemize}
  \item \ruleref{\ottdruleDTVarLabel}
    says that the type of a variable is inferred from the context.
  \item \ruleref{\ottdruleDTThunkLabel} says
    that the type of a thunk is inferred by shifting up the type of the 
    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
    infers the type of a return by shifting down the type of the
    contained value.
  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
    They allow the inferred type to be refined by annotating it with a supertype.
  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
    mean that the declarative system allows us to infer any type from the equivalence class.
  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
    and its first premise infers the existential type of the value being unpacked.
    It is important however that the inferred existential type is normalized. 
    This is because there might be multiple equivalent existential types 
    with a different order or even number of quantified variables, 
    and to bind them, the algorithm needs to fix the canonical one.
  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
    with annotated let-bindings. The first premise infers the type of the head of the application,
    which must be a thunked computation. Then if after applying it 
    to the arguments, the resulting type can be equated to the annotated one,
    we infer the body of the let-binding in the context extended with the bound variable.
  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
    bus is used for unannotated let-bindings. In this case, 
    we require the type application to infer the `canonical' principal type.
    $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
    any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
    is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.
\end{itemize}

@@

Let us discuss the inference rules of the algorithm:
\begin{itemize}
  \item \ruleref{\ottdruleATVarLabel} 
    infers the type of a variable by looking it up in the context
    and normalizing the result.
  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
    are similar to the declarative rules: they make a recursive call
    to type the body of the thunk or the return expression and
    put the shift on top of the result.
  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
    are symmetric. They make a recursive call to infer the type of the annotated
    expression, check that the inferred type is a subtype of the annotation,
    and return the normalized annotation.
  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
    It makes a recursive call to infer the type of the body in the extended context,
    and returns the corresponding arrow type. Notice that the algorithm also
    normalizes the result, which is because the annotation type $[[iP]]$
    is allowed to be non-normalized.
  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
    Similarly to the previous case, it makes a recursive call to infer the type
    of the body in the extended \emph{type} context. 
    After that, it returns the corresponding universal type. 
    It is also required to normalize the result, because, 
    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
    in which case the $[[∀]]$ must be removed. 
  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
    it makes a recursive call to infer the type of the bound value,
    and then returns the type of the body in the extended context.
  \item \ruleref{\ottdruleATAppLetAnnLabel}
    is interpreted as follows.
    First, it infers the type of the head of the application,
    ensuring that it is a thunked computation $[[↓iM]]$;
    after that, it makes a recursive call
    to the application inference procedure,
    which returns the algorithmic type, whose
    instantiation to a declarative type must be associated with the bound variable 
    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
    check whether the instantiation to the annotated type $[[iP]]$ is possible,
    and if it is, the algorithm infers the type of the body in the extended context,
    and returns it as the result. 
  \item \ruleref{\ottdruleATAppLetLabel}
    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
    However, since there is no annotation to assign the result to,
    the algorithm must infer the `canonical' principal type.
    To do that algorithmically, we ensure that 
    the inferred algorithmic type $[[uQ]]$ is instantiated
    to the minimal possible type $[[ [uσ]uQ ]]$.
    The premise $[[Γ ⊢ uQ SC minby uσ]]$
    provides the minimal instantiation of $[[uQ]]$ \wrt $[[SC]]$. It 
    guarantees that if we consider all possible substitutions satisfying the inferred 
    constraints $[[SC]]$, then substitution $[[uσ]]$ will instantiate
    $[[uQ]]$ to the minimal possible type $[[ [uσ]uQ ]]$.
    This will be the principal type that we assign to the result of the application 
    (bound to the variable $[[x]]$)
    and then we infer the type of the body in the context extended with 
    the bound variable $[[x : [uσ]uQ]]$.
  \item \ruleref{\ottdruleATUnpackLabel}
    first, infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
    and since the type is guaranteed to be normalized, binds 
    the quantified variables with $[[nas]]$.
    Then it infers the type of the body in the appropriately extended context
    and checks that the inferred type does not depend on $[[nas]]$
    by checking well-formedness $[[Γ ⊢ iN]]$.
\end{itemize}

@@

The correctness of the algorithm consists of its soundness and 
completeness, which is by mutual
induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
The simplified result is the following.
\begin{theorempreview}
  \hfill
  \begin{itemize}
    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
  \end{itemize}
\end{theorempreview}

@@

@@

>>>>

\section{Relation to System F}

The types and terms of \fexists have an additional polarization structure.
Despite that, it is closely related to the well-known \systemf.
To demonstrate that, we provide two translations in both ways:
the polarization from \systemf to \fexists and the depolarization from 
\fexists to \systemf. These translations are done at the level of types and terms,
and the expected typing preservation properties are proved.

\begin{figure}[h]
  \ottdefnSFtyLabeled{}
  \ottdefnSFtyALabeled{}
  \caption{Typing rules of \systemf}
  \label{fig:systemf-typing}
\end{figure}

\begin{figure}[h]
  \ottdefnSFWFLabeled{}
\end{figure}

\subsection{Type-level Translation}

\begin{figure}[h]
  \ottfundefnunpolP{}
  \ottfundefnunpolN{}
\end{figure}

\begin{figure}[h]
  \ottdefnELNLabeled{}
  \ottdefnELPLabeled{}
  \caption{Subtyping elaboration from \fexists to \systemf}
  \label{fig:subt-elaboration}
\end{figure}

% \lemmaSubtElabSoundness

% \lemmaElabRemovable

% \lemmaDepolSubst

\subsection{Term-level Translation}

The term-level translation is defined not solely for terms, but for typing
derivations. The reason for that is that the terms of \systemf and \fexists
contain different typing information that cannot be reconstructed without the
appropriate derivation tree. For instance, the lambda expressions in \systemf do
not have the bound variable annotation, while the lambda expressions in \fexists
do; on the other hand, \fexists has rich subtyping, which cannot be expressed in
in terms of polymorphic type instantiation of \systemf.

\begin{figure}[h]
  \ottdefnTELPLabeled{}
  \caption{Typing elaboration from \fexists to \systemf}
  \label{fig:typing-elaboration}
\end{figure}

\begin{figure}[h]
  \ottdefnTELNLabeled{}
\end{figure}

\begin{figure}[h]
  \ottdefnTELAppLabeled{}
\end{figure}

\begin{figure}[h]
  \ottdefnSFfPolLabeled{}
  \caption{Polarization: translation from \systemf to \fexists}
  \label{fig:systemf-polarization}
\end{figure}

% \lemmaDepolarSoundness*

% \lemmaPolSubst

% \lemmaPolRemovable

% \lemmaTypPolWf*

% \lemmaPolarSoundness

@@

Let us discuss the selected rules of the declarative system:
\begin{itemize}
  \item \ruleref{\ottdruleDTVarLabel}
    says that the type of a variable is inferred from the context.
  \item \ruleref{\ottdruleDTThunkLabel} says
    that the type of a thunk is inferred by shifting up the type of the 
    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
    infers the type of a return by shifting down the type of the
    contained value.
  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
    They allow the inferred type to be refined by annotating it with a supertype.
  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
    mean that the declarative system allows us to infer any type from the equivalence class.
  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
    and its first premise infers the existential type of the value being unpacked.
    It is important however that the inferred existential type is normalized. 
    This is because there might be multiple equivalent existential types 
    with a different order or even number of quantified variables, 
    and to bind them, the algorithm needs to fix the canonical one.
  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to
    type the \emph{annotated} applicative let-binders. 
    The first premise infers the type of the head of the application,
    which must be a thunked computation. Then if after applying it 
    to the arguments, the resulting type can be instantiated to the annotated one,
    we infer the body of the let-binding in the context extended with the bound variable.
  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
    bus is used for unannotated let-bindings. In this case, 
    we require the type application to infer the `canonical' principal type.
    $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
    any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
    is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.
\end{itemize}

@@

Let us discuss the inference rules of the algorithm:
\begin{itemize}
  \item \ruleref{\ottdruleATVarLabel} 
    infers the type of a variable by looking it up in the context
    and normalizing the result.
  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
    are similar to the declarative rules: they make a recursive call
    to type the body of the thunk or the return expression and
    put the shift on top of the result.
  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
    are symmetric. They make a recursive call to infer the type of the annotated
    expression, check that the inferred type is a subtype of the annotation,
    and return the normalized annotation.
  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
    It makes a recursive call to infer the type of the body in the extended context,
    and returns the corresponding arrow type. Notice that the algorithm also
    normalizes the result, which is because the annotation type $[[iP]]$
    is allowed to be non-normalized.
  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
    Similarly to the previous case, it makes a recursive call to infer the type
    of the body in the extended \emph{type} context. 
    After that, it returns the corresponding universal type. 
    It is also required to normalize the result, because, 
    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
    in which case the $[[∀]]$ must be removed. 
  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
    it makes a recursive call to infer the type of the bound value,
    and then returns the type of the body in the extended context.
  \item \ruleref{\ottdruleATAppLetAnnLabel}
    is interpreted as follows.
    First, it infers the type of the head of the application,
    ensuring that it is a thunked computation $[[↓iM]]$;
    after that, it makes a recursive call
    to the application inference procedure,
    which returns the algorithmic type, whose
    instantiation to a declarative type must be associated with the bound variable 
    $[[x]]$; then premise $[[Γ; Θ ⊨ uM' ≤ ↑iP ⫤ SC2]]$
    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
    check whether the instantiation to the annotated type $[[iP]]$ is possible,
    and if it is, the algorithm infers the type of the body in the extended context,
    and returns it as the result. 
  \item \ruleref{\ottdruleATAppLetLabel}
    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
    However, since there is no annotation to assign the result to,
    the algorithm must infer the `canonical' principal type.
    To do that algorithmically, we ensure that 
    the inferred algorithmic type $[[uQ]]$ is instantiated
    to the minimal possible type $[[ [uσ]uQ ]]$.
    The premise $[[Γ ⊢ uQ SC minby uσ]]$
    provides the minimal instantiation of $[[uQ]]$ \wrt $[[SC]]$. It 
    guarantees that if we consider all possible substitutions satisfying the inferred 
    constraints $[[SC]]$, then substitution $[[uσ]]$ will instantiate
    $[[uQ]]$ to the minimal possible type $[[ [uσ]uQ ]]$.
    This will be the principal type that we assign to the result of the application 
    (bound to the variable $[[x]]$)
    and then we infer the type of the body in the context extended with 
    the bound variable $[[x : [uσ]uQ]]$.
  \item \ruleref{\ottdruleATUnpackLabel}
    first, infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
    and since the type is guaranteed to be normalized, binds 
    the quantified variables with $[[nas]]$.
    Then it infers the type of the body in the appropriately extended context
    and checks that the inferred type does not depend on $[[nas]]$
    by checking well-formedness $[[Γ ⊢ iN]]$.
\end{itemize}

@@

The correctness of the algorithm consists of its soundness and 
completeness, which is proved by mutual
induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
The simplified result is the following.
\begin{theorempreview}
  \hfill
  \begin{itemize}
    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
  \end{itemize}
\end{theorempreview}

@@

\section{Relation to \systemf}
\input{\genDir/tex/translation-statements.tex}

@@

\section{Relation to \systemf}
\input{\genDir/tex/translation-proofs.tex}

### 2024-03-15 proofs/tex/decl-typing-lemmas.tex
wip

        \item \ruleref{\ottdruleDTVarLetLabel}
            Then we prove that 
            $[[Γ ; Φ1 ⊢ let x = v; c : iN]]$ implies $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$.
            First, we apply the induction hypothesis to 
            $[[Γ; Φ1 ⊢ v : iP]]$ to obtain $[[Γ; Φ2 ⊢ v : iP]]$ 
            of the same pure size.

@@

@@

            As in the previous case, we apply the induction hypothesis to each of the premises
            and rewrite:
            \begin{itemize}
                \item $[[Γ ; Φ1 ⊢ v : ↓iM]]$ into $[[Γ ; Φ2 ⊢ v : ↓iM]]$,
                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ]]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$, 
                    and
                \item $[[Γ ; Φ1, x:iP ⊢ c : iN]]$ into $[[Γ ; Φ2, x:iP ⊢ c : iN]]$
                (notice that $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$).
            \end{itemize}

            Notice that $[[Γ ⊢ iP]]$ and $[[Γ ⊢ ↑iQ ≤ ↑iP]]$ 
            do not depend on the variable context, and hold by assumption.
            Then we infer $[[Γ ; Φ2 ⊢ let x:iP = v(args); c : iN]]$ by \ruleref{\ottdruleDTAppLetAnnLabel}.

>>>>

        \item \ruleref{\ottdruleDTVarLetLabel}
            Then we prove that 
            $[[Γ ; Φ1 ⊢ let x = v; c : iN]]$ implies $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$.
            First, we apply the induction hypothesis to 
            $[[Γ; Φ1 ⊢ v : iP]]$ to obtain $[[Γ; Φ2 ⊢ v : iP]]$ 
            of the same size.

@@

        \item \ruleref{\ottdruleDTCVarLetLabel}
            Then we prove that 
            $[[Γ ; Φ1 ⊢ let x : iP = c; c' : iN]]$ implies $[[Γ ; Φ2 ⊢ let x : iP = c; c' : iN]]$.

            First, we apply the induction hypothesis to 
            $[[Γ; Φ1 ⊢ c : iM]]$ to obtain $[[Γ; Φ2 ⊢ c : iM]]$
            of the same size.

            Then we apply the induction hypothesis to
            the equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
            and the premise $[[Γ ; Φ1, x:iP ⊢ c' : iN]]$ to obtain
            $[[Γ ; Φ2, x:iP ⊢ c' : iN]]$.
            Then we infer $[[Γ ; Φ2 ⊢ let x : iP = c; c' : iN]]$ by \ruleref{\ottdruleDTCVarLetLabel}.

@@

            As in the previous case, we apply the induction hypothesis to each of the premises
            and rewrite:
            \begin{itemize}
                \item $[[Γ ; Φ1 ⊢ v : ↓iM]]$ into $[[Γ ; Φ2 ⊢ v : ↓iM]]$,
                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> iM']]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> iM']]$, 
                    and
                \item $[[Γ ; Φ1, x:iP ⊢ c : iN]]$ into $[[Γ ; Φ2, x:iP ⊢ c : iN]]$
                (notice that $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$).
            \end{itemize}

            Notice that $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM' ≤ ↑iP]]$ 
            do not depend on the variable context, and hold by assumption.
            Then we infer $[[Γ ; Φ2 ⊢ let x:iP = v(args); c : iN]]$ by \ruleref{\ottdruleDTAppLetAnnLabel}.
